A Gentle Introduction and Application of Feature-Based Clustering with Psychological Time Series

### 2023-02-24 latex-results/05FeatureClustering.tex
:construction: cluster explanation

For our illustration of the feature clustering procedure, we use the
generalizable and efficient centroid-based k-means clustering. The
procedure needs the user to make a selection on the optimal number of
clusters in order to run the main algorithm. We use the Elbow and the
Silhouette method, which both suggest two clusters to be the optimal
solution (see Supplemental Material A). We then entered the
participants' PC-scores into the k-means algorithm. We used the
recommended Hartigan and Wong algorithm \citep{hartigan1979} with 100
random initial centroid positions to avoid convergence to a sub-optimal
solution (i.e., local minima). The k-means analysis assigned 80
participants to cluster 1 and 76 participants to cluster 2.

>>>>

For the actual k-means clustering, the analysis seeks to minimize the
total within-cluster variation. We used the Hartigan and Wong algorithm,
which is a widely used approach in k-means clustering of time series
features \citep{hartigan1979}. This algorithm is designed to optimize
the clustering of the feature data into \(k\) groups, where \(k\) is a
pre-defined number of clusters. The algorithm starts by randomly
separating the data points into k clusters and then iteratively updates
the assignment of each point to the nearest cluster center until
convergence. The Hartigan and Wong algorithm specifically calculates the
within-cluster variation (\(W\)) of cluster \(C_i\) as the summed
squared Euclidean distances of the feature \(x\) to its assigned cluster
centroid \(\mu_i\):

\begin{equation} \label{eq:kWCi}
  W(C_i) = \sum_{x \in C_i}(x-\mu_i)^2
\end{equation}

By summing the within cluster sum of squares from all \(k\) clusters, we
can then derive the total within-cluster sum of square \(WCSS\):

\begin{equation} \label{eq:kWCSS}
  WCSS = \sum_{i=1}^k W(C_i) = \sum_{i=1}^k \sum_{x \in C_i} (x - \mu_i)^2
\end{equation}

It is this \(WCSS\) that becomes the objective function to be minimized,
by iteratively moving features from one cluster to another
\citep{hartigan1979}. In particular, the algorithm (1) calculates the
cluster centroids of the initial partitioning, (2) checks whether any
feature has a centroid that is closer than that of the currently
assigned cluster (3) updates the centroids based on any reassigned
features, and then iterates between steps two and three until \(WCSS\)
is minimized (i.e., locally optimal convergence) or a maximum number of
iterations is reached. Given the iterative nature of the algorithm, the
initial partitioning is often important because the algorithm might
arrive at a suboptimal clustering where the \(WCSS\) cannot be further
reduced by moving any feature to another cluster, despite a better
solution existing (i.e., a local minimum). It is, therefore often
recommended to run the k-means clustering with several different
starting positions.

In our case we used the Elbow and the Silhouette method to determine the
recommended number of \(k\) clusters, which for both methods suggested
two clusters (see Supplemental Material A). We then entered the
participants' PC-scores from the previous step into the k-means
algorithm. To avoid local minima we used 100 random initial centroid
positions. In the final solution the k-means analysis assigned 80
participants to cluster 1 and 76 participants to cluster 2.

### 2023-02-24 sections/05FeatureClustering.Rmd
:construction: cluster explanation

For our illustration of the feature clustering procedure, we use the generalizable and efficient centroid-based k-means clustering. The procedure needs the user to make a selection on the optimal number of clusters in order to run the main algorithm. We use the Elbow and the Silhouette method, which both suggest two clusters to be the optimal solution (see Supplemental Material A). We then entered the participants' PC-scores into the k-means algorithm. We used the recommended Hartigan and Wong algorithm \citep{hartigan1979} with 100 random initial centroid positions to avoid convergence to a sub-optimal solution (i.e., local minima). The k-means analysis assigned `r table(kmeansOut$cluster)[1]` participants to cluster 1 and `r table(kmeansOut$cluster)[2]` participants to cluster 2.

>>>>

For the actual k-means clustering, the analysis seeks to minimize the total within-cluster variation. We used the Hartigan and Wong algorithm, which is a widely used approach in k-means clustering of time series features \citep{hartigan1979}. This algorithm is designed to optimize the clustering of the feature data into $k$ groups, where $k$ is a pre-defined number of clusters. The algorithm starts by randomly separating the data points into k clusters and then iteratively updates the assignment of each point to the nearest cluster center until convergence. The Hartigan and Wong algorithm specifically calculates the within-cluster variation ($W$) of cluster $C_i$ as the summed squared Euclidean distances of the feature $x$ to its assigned cluster centroid $\mu_i$:

```{=tex}
\begin{equation} \label{eq:kWCi}
  W(C_i) = \sum_{x \in C_i}(x-\mu_i)^2
\end{equation}
```

By summing the within cluster sum of squares from all $k$ clusters, we can then derive the total within-cluster sum of square $WCSS$:

```{=tex}
\begin{equation} \label{eq:kWCSS}
  WCSS = \sum_{i=1}^k W(C_i) = \sum_{i=1}^k \sum_{x \in C_i} (x - \mu_i)^2
\end{equation}
```

It is this $WCSS$ that becomes the objective function to be minimized, by iteratively moving features from one cluster to another \citep{hartigan1979}. In particular, the algorithm (1) calculates the cluster centroids of the initial partitioning, (2) checks whether any feature has a centroid that is closer than that of the currently assigned cluster (3) updates the centroids based on any reassigned features, and then iterates between steps two and three until $WCSS$ is minimized (i.e., locally optimal convergence) or a maximum number of iterations is reached. Given the iterative nature of the algorithm, the initial partitioning is often important because the algorithm might arrive at a suboptimal clustering where the $WCSS$ cannot be further reduced by moving any feature to another cluster, despite a better solution existing (i.e., a local minimum). It is, therefore often recommended to run the k-means clustering with several different starting positions.

In our case we used the Elbow and the Silhouette method to determine the recommended number of $k$ clusters, which for both methods suggested two clusters (see Supplemental Material A). We then entered the participants' PC-scores from the previous step into the k-means algorithm. To avoid local minima we used 100 random initial centroid positions. In the final solution the k-means analysis assigned `r table(kmeansOut$cluster)[1]` participants to cluster 1 and `r table(kmeansOut$cluster)[2]` participants to cluster 2.
